Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2020-11-24 sources/bmc_template_submission/bmc_article.tex
bmc discussion citations

Phylogenetic pipelines designed to make evolutionary sense of the vast amount of public molecular data (e.g., Phylota [@sanderson2008phylota], PHLAWD [@smith2009mega], SUPERSMART [@antonelli2017toward]) focus on generating full phylogenies *de novo*, i.e., inferring phylogenetic relationships from a newly generated homology hypothesis, as opposed to e.g., supertrees, that are generated by summarizing previous phylogenetic estimates.
While Physcraper does not generate phylogenies *de novo* in a traditional sense,
it successfully generates new phylogenetic knowledge, revealing the importance of
open science in facilitating phylogenetic placement of public molecular data and accelerating enrichment and updating of phylogenetic relationships in any region of the tree of life.
The PUMPER pipeline [@izquierdo2014pumper] also uses the concept of updating
pre-existing alignments to incorporate public molecular data into phylogenies. Unfortunately, installation was unsuccessful following instructions from the author, and a comparison analysis was unfeasible.
<!-- **MTH: add some more details about what aspect of the PUMPER install failed** -->

Physcraper generates individual gene trees, failing to capture the complexity of species' evolutionary history [@song2012resolving]. Yet, Physcraper facilitates gathering alignments and gene trees for multiple loci from a group of interest, that can be used to reconstruct species trees with ASTRAL [@mirarab2014astral], BEAST2 [@bouckaert2019beast], or SVD Quartets [@chifman2014quartet]).

Physcraper can potentially link phylogenies to data available in any of the taxonomies integrated in the OpenTree taxonomy [@rees2017automated], such as geographical locations from the Global Biodiversity Information Facility, or fossils from the Paleobiology Database.
<!-- a resource that is contributing to achieve the goal of "exploring and analysing biodiversity at an accelerated pace, and returning systematics into the mainstream of science" [@wilson2003encyclopedia]. -->
The Physcraper workflow can be used to rapidly (in a matter of hours)
address challenges overarching both fields of ecology and evolution, such as
placing newly discovered species phylogenetically [@webb2010biodiversity],
systematizing molecular (and other) databases, i.e., curating taxonomic assignations [@san2010molecular],
and generating custom trees for ecological [@helmus2012phylogenetic] and evolutionary downstream analyses [@stoltzfus2013phylotastic].

Data repositories hold more information than meets the eye.
Beyond the main data, they are rich sources of metadata that can be leveraged for the advantage of all areas of biology as well as the advancement of scientific policy and applications.
Initial ideas about the data are constantly changed by results from new analyses.
Physcraper provides a framework for reproducible phylogenetics that has the potential to consistently provide context for these ideas, highlighting the importance of data sharing and open science in the field, biology and science.
\section*{Conclusions}
Text for this section\ldots

>>>>

Phylogenetic pipelines designed to make evolutionary sense of the vast amount of
public molecular data (e.g., Phylota \cite{sanderson2008phylota},
PHLAWD \cite{smith2009mega},
SUPERSMART \cite{antonelli2017toward}) focus on generating full phylogenies *de novo*,
i.e., inferring phylogenetic relationships from a newly generated homology hypothesis,
as opposed to e.g., supertrees, that are generated by summarizing previous
phylogenetic estimates.
While Physcraper does not generate phylogenies *de novo* in a traditional sense,
it successfully generates new phylogenetic knowledge, revealing the importance of
open science in facilitating phylogenetic placement of public molecular data and
accelerating enrichment and updating of phylogenetic relationships in any region
of the tree of life.
The PUMPER pipeline \cite{izquierdo2014pumper} also uses the concept of updating
pre-existing alignments to incorporate public molecular data into phylogenies.
Unfortunately, installation was unsuccessful following instructions from the author,
and a comparison analysis between Physcraper and PUMPER is unfeasible at present.
%% **MTH: add some more details about what aspect of the PUMPER install failed**

Physcraper generates individual gene trees, failing to capture the complexity of
species' evolutionary history \cite{song2012resolving}. Yet, Physcraper facilitates
gathering alignments and gene trees for multiple loci from a group of interest,
that can be used to reconstruct species trees with ASTRAL \cite{mirarab2014astral},
BEAST2 \cite{bouckaert2019beast}, or SVD Quartets \cite{chifman2014quartet}).

Physcraper can potentially link phylogenies to data available in any of the
taxonomies integrated in the OpenTree taxonomy \cite{rees2017automated}, such as
geographical locations from the Global Biodiversity Information Facility, or
fossils from the Paleobiology Database.
%% a resource that is contributing to achieve the goal of "exploring and
%% analysing biodiversity at an accelerated pace, and returning systematics into
%% the mainstream of science" \cite{wilson2003encyclopedia}.
The Physcraper workflow can be used to rapidly (in a matter of hours)
address challenges overarching both fields of ecology and evolution, such as
phylogenetically placing newly discovered species \cite{webb2010biodiversity},
systematizing molecular (and other) databases, i.e., curating taxonomic assignations
\cite{san2010molecular},
and generating custom trees for ecological \cite{helmus2012phylogenetic} and
evolutionary downstream analyses \cite{stoltzfus2013phylotastic}.

Data repositories hold more information than meets the eye.
Beyond the main data, they are rich sources of metadata that can be leveraged
for the advantage of all areas of biology as well as the advancement of scientific
policy and applications.
Initial ideas about the data are constantly changed by results from new analyses.
Physcraper provides a framework for reproducible phylogenetics that has the
potential to consistently provide context for these ideas, highlighting the
importance of data sharing and open science in the field, biology and science.

\section*{Conclusions}
Text for this section\ldots
